---
title: "Project4"
author: "Jin Ah Kang"
date: "5/10/2019"
output: html_document
---

#1.Use this piece of code to download and prepare data for use in project
```{r 1}
library(tidyverse)
library(stringr)

arrest_tab <- read_csv("http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv")

dat <- arrest_tab %>% 
  filter(!is.na(`Location 1`)) %>%
  separate(`Location 1`, c("lat","lng"), sep=",") %>%
  mutate(lat=as.numeric(str_replace(lat, "\\(", ""))) %>%
  mutate(lng=as.numeric(str_replace(lng, "\\)", ""))) %>%
  sample_n(2000)
dat
```

#2.Use the leaflet package to create an interactive map of Baltimore
```{r 2}
library(leaflet)

balto_map <- leaflet(dat) %>%
  addTiles() %>%
  setView(lat=39.29, lng=-76.61, zoom=11)
balto_map
```

#3.Add graphical elements to display the data. For instance, add circles, with colors indicating sex. Or circles with colors indicating race. Or anything else that strikes your fancy.
```{r 3 first}
getColor <- function(dat) {
  sapply(dat$age, function(age) {
    if(age < 20) {
      "red"
    } else if(age < 30) {
      "white"
    } else if(age < 40) {
      "orange"
    } else if(age < 50) {
      "blue"
    } else if(age < 60) {
      "green"
    } else {
      "black"
    }
    })
}

icons <- awesomeIcons(
  icon = 'fa-user',
  library = 'fa',
  markerColor = getdColor(dat)
)

balto_map %>% addAwesomeMarkers(icon = icons, label = ~as.character(age))
```

#(a) code to carry out each of the steps above
#(b) output showing the result of your code
#(c) a short prose description of your interactive map
I divide the marks according to ages; The mark in the map shows that there are more white mark (age between 20 and 29) than any other age groups; the orange mark (age between 30 and 39) and the green mark (age between 40 and 49) are shown equally as many; one green mark (age between 50 and 59) is shown among as many as 5 white, blue or orange marks and most of the green mark appears towards center (the further away from the center of Baltimore, the less green mark we can observe); the black mark (age of 60 or more) is very rare that it can even be counted with by observing the map. 

```{r analysis}
a0 <- dat %>%
  filter(age < 20) %>% count()
a1 <- dat %>%
  filter(age >= 20 & age < 30) %>% count()
a2 <- dat %>%
  filter(age >= 30 & age < 40) %>% count()
a3 <- dat %>%
  filter(age >= 40 & age < 50) %>% count()
a4 <- dat %>%
  filter(age >= 50 & age < 60) %>% count()
a5 <- dat %>%
  filter(age >= 60) %>% count()

a0
a1
a2
a3
a4
a5
```